---
title: "Getting Started with shinyds"
output: html
vignette: >
  %\VignetteIndexEntry{Getting Started with shinyds}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(shinyds)
```

## Introduction

`shinyds` provides R wrappers for [Designsystemet](https://designsystemet.no), the Norwegian
government's design system. It lets you build Shiny applications that follow Norwegian public
sector design guidelines using familiar `ds_*` functions.

Components come from two upstream sources:

- **CSS components** — plain HTML elements styled with `ds-*` CSS classes (`ds_button()`,
  `ds_alert()`, `ds_input()`, …)
- **Web components** — custom HTML elements with built-in JavaScript behaviour (`ds_tabs()`,
  `ds_pagination()`, `ds_suggestion()`)

## Installation

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("your-org/shinyds")
```

## Minimal app

Every app needs `use_designsystemet()` in the UI. It loads the CSS and JavaScript bundles and
activates the design token colour scheme.

```{r}
#| eval: false
library(shiny)
library(bslib)
library(shinyds)

ui <- bslib::page_fluid(
  use_designsystemet(),

  ds_heading("Hello Designsystemet!", level = 1),
  ds_paragraph("A Shiny app using Norwegian government design components."),

  ds_field(
    ds_label("Your name", `for` = "name"),
    ds_input("name", placeholder = "Enter your name")
  ),

  ds_button("Submit", inputId = "submit", variant = "primary"),

  verbatimTextOutput("out")
)

server <- function(input, output, session) {
  output$out <- renderPrint(list(name = input$name, clicks = input$submit))
}

shinyApp(ui, server)
```

`bslib::page_fluid()` is the recommended page function — it provides Bootstrap 5 and works
cleanly alongside Designsystemet's own CSS tokens.

## Form controls

### Button

```{r}
#| eval: false
# label first, then inputId
ds_button("Click me",  inputId = "btn1", variant = "primary")
ds_button("Secondary", inputId = "btn2", variant = "secondary")
ds_button("Tertiary",  inputId = "btn3", variant = "tertiary")

# sizes
ds_button("Small",  size = "sm")
ds_button("Medium", size = "md")
ds_button("Large",  size = "lg")

# states
ds_button("Disabled", variant = "primary", disabled = TRUE)
ds_button("Loading",  variant = "secondary", loading = TRUE)
```

`input$btn1` starts at `0` and increments by 1 on each click.

### Text inputs

```{r}
#| eval: false
ds_field(
  ds_label("Email", `for` = "email"),
  ds_input("email", type = "email", placeholder = "you@example.com"),
  ds_validation_message("Must be a valid email address", variant = "error")
)

ds_field(
  ds_label("Message", `for` = "msg"),
  ds_textarea("msg", placeholder = "Type here…", rows = 4)
)
```

Wrap `ds_input()` / `ds_textarea()` in `ds_field()` with a `ds_label()` — this is the standard
Designsystemet pattern and ensures correct label association and spacing.

### Checkbox and radio

```{r}
#| eval: false
ds_checkbox("agree", label = "I accept the terms")

# Radio buttons — group them with a shared name
ds_radio("opt_a", label = "Option A", value = "a", name = "opts", checked = TRUE)
ds_radio("opt_b", label = "Option B", value = "b", name = "opts")
```

`input$agree` returns `TRUE` / `FALSE`. Radio inputs are individual elements; read any one of the
group's `inputId` values to get the selected `value`.

### Select

```{r}
#| eval: false
ds_field(
  ds_label("Country", `for` = "country"),
  ds_select("country",
    choices  = c("Norway" = "no", "Sweden" = "se", "Denmark" = "dk"),
    selected = "no"
  )
)
```

`input$country` returns the selected option's value string.

### Fieldset

Group related controls under a shared legend:

```{r}
#| eval: false
ds_fieldset(
  legend = "Notification preferences",
  ds_checkbox("notif_email", label = "Email"),
  ds_checkbox("notif_sms",   label = "SMS"),
  ds_checkbox("notif_push",  label = "Push")
)
```

> **Note:** `ds_fieldset()` is backed by a behaviour-only JavaScript module. See the
> *Reactivity Patterns* vignette if you need to react to fieldset-level events.

### Search and suggestion

```{r}
#| eval: false
# Search input — input$q returns the current text
ds_field(
  ds_label("Search", `for` = "q"),
  ds_search("q", placeholder = "Search…")
)

# Autocomplete suggestion — input$fruit returns the selected value
ds_field(
  ds_label("Fruit", `for` = "fruit"),
  ds_suggestion("fruit",
    choices     = c("Apple", "Banana", "Cherry"),
    placeholder = "Start typing…"
  )
)
```

## Typography

```{r}
#| eval: false
# Headings — level sets the HTML element (h1–h6), size sets the visual token
ds_heading("Page Title",    level = 1, size = "2xl")
ds_heading("Section Title", level = 2, size = "lg")
ds_heading("Card Title",    level = 3, size = "md")

ds_paragraph("Body copy.", size = "md")
ds_paragraph("Small caption.", size = "sm")

ds_link("Designsystemet", href = "https://designsystemet.no")

ds_list(
  ds_list_item("First"),
  ds_list_item("Second"),
  ds_list_item("Third"),
  ordered = TRUE
)
```

## Layout

### Cards

```{r}
#| eval: false
ds_card(
  ds_card_block(
    ds_heading("Card Title", level = 3, size = "sm"),
    ds_paragraph("Card content.")
  )
)

ds_card(variant = "tinted",
  ds_card_block("Highlighted content")
)
```

### Tables

```{r}
#| eval: false
ds_table(
  ds_thead(
    ds_tr(ds_th("Name"), ds_th("Role"), ds_th("Status"))
  ),
  ds_tbody(
    ds_tr(ds_td("Alice"), ds_td("Developer"),
          ds_td(ds_tag("Active", color = "success"))),
    ds_tr(ds_td("Bob"),   ds_td("Designer"),
          ds_td(ds_tag("Active", color = "success")))
  )
)
```

### Tabs (web component)

```{r}
#| eval: false
ds_tabs("my_tabs",
  ds_tablist(
    ds_tab("Overview",  value = "overview",  selected = TRUE),
    ds_tab("Details",   value = "details")
  ),
  ds_tabpanel(value = "overview",
    ds_paragraph("Overview content.")
  ),
  ds_tabpanel(value = "details",
    ds_paragraph("Details content.")
  )
)

# In server: input$my_tabs returns the selected tab value string
```

### Pagination (web component)

```{r}
#| eval: false
ds_pagination("pager", current = 1, total = 10)

# In server: input$pager returns the current page number (integer)
```

## Feedback components

```{r}
#| eval: false
ds_alert("Informational message.", variant = "info")
ds_alert("Operation succeeded.",   variant = "success")
ds_alert("Review before saving.",  variant = "warning")
ds_alert("Something went wrong.",  variant = "danger")

ds_spinner(title = "Loading…", size = "md")

ds_skeleton(variant = "text",      width = "100%")
ds_skeleton(variant = "circle",    width = "48px", height = "48px")
ds_skeleton(variant = "rectangle", width = "200px", height = "80px")

ds_badge_position(
  ds_button("Inbox", variant = "secondary"),
  ds_badge(count = 4, color = "danger")
)
```

## Updating inputs from the server

Several inputs support programmatic updates:

```{r}
#| eval: false
server <- function(input, output, session) {
  observeEvent(input$reset, {
    update_ds_input(session,    "name",    value = "")
    update_ds_checkbox(session, "agree",   value = FALSE)
    update_ds_select(session,   "country", value = "no")
    update_ds_tabs(session,     "my_tabs", selected = "overview")
    update_ds_pagination(session, "pager", current = 1)
  })
}
```

## Using bslib layout primitives

Because `shinyds` components are plain HTML tags, they compose naturally with `bslib` layout
functions:

```{r}
#| eval: false
ui <- bslib::page_fluid(
  use_designsystemet(),

  bslib::layout_sidebar(
    sidebar = bslib::sidebar(
      title    = "Controls",
      position = "right",
      width    = 280,
      verbatimTextOutput("values")
    ),

    # Main content
    ds_tabs("tabs",
      ds_tablist(
        ds_tab("Form",   value = "form",   selected = TRUE),
        ds_tab("Result", value = "result")
      ),
      ds_tabpanel(value = "form",
        ds_field(ds_label("Name", `for` = "nm"), ds_input("nm")),
        ds_button("Submit", inputId = "go")
      ),
      ds_tabpanel(value = "result",
        verbatimTextOutput("out")
      )
    )
  )
)
```

Use `bslib::layout_columns()` for equal-width column grids, and
`bslib::layout_column_wrap(width = "220px")` for auto-fitting responsive grids.

## Running the example apps

```{r}
#| eval: false
# Minimal form example
shiny::runApp(system.file("examples/basic", package = "shinyds"))

# Data visualisation with tabs and layout_sidebar
shiny::runApp(system.file("examples/faithful", package = "shinyds"))

# Full component reference
shiny::runApp(system.file("examples/showcase", package = "shinyds"))
```

## Next steps

- **Reactivity Patterns** — how to add Shiny reactivity to behaviour-only module components
  (toggle group, details, dialog, popover, fieldset):
  `vignette("reactivity-patterns", package = "shinyds")`
- [Designsystemet documentation](https://designsystemet.no)
- [Shiny documentation](https://shiny.posit.co)
